---
title: "Lab 06 - Ugly charts and Simpson's paradox"
author: "Noah Booker"
date: "3/28/25"
output: github_document
---
# Take a Sad Plot and Make It Better

## Instructional staff employment trends
### Load packages and data

```{r load-packages, message = FALSE}
library(tidyverse) 
library(scales)
#library(dsbox)
#library(mosaicData) 
```


```{r codefromlabintro}
staff <- read_csv("data/instructional-staff.csv")

staff_long <- staff %>%
  pivot_longer(cols = -faculty_type, names_to = "year") %>%
  mutate(value = as.numeric(value))

staff_long %>%
  ggplot(aes(x = year, y = value, color = faculty_type)) +
  geom_line()

staff_long %>%
  ggplot(aes(
    x = year,
    y = value,
    group = faculty_type,
    color = faculty_type
  )) +
  geom_line()
```

### Exercise 1

Include the line plot you made above in your report and make sure the figure width is large enough to make it legible.
Also fix the title, axis labels, and legend label.

```{r plot1, fig.width=10, fig.height=6}
staff_long %>%
  ggplot(aes(
    x = year,
    y = value,
    group = faculty_type,
    color = faculty_type
  )) +
  geom_line() +
  labs(title = "Instructional Staff Employees, 1975-2011",
       x = "Year", y = "Percentage of Hires",
       color = "Faculty Type") +
  scale_y_continuous(labels = label_percent(scale = 1)) +
  theme_minimal()
```


### Exercise 2

Suppose the objective of this plot was to show that the proportion of part-time faculty have gone up over time compared to other instructional staff types. What changes would you propose making to this plot to tell this story?

The current plot seems to tell that story pretty well. The only line that shows a consistent positive slope over the years is Part-Time Faculty. But perhaps I'm overlooking some changes that could make that fact more obvious...

Solution: I had the idea of trying to make the line for Part-Time Faculty thicker than the other lines, and asked Claude.ai for a way to do this. Claude helped me to create the code below which adds in a separate geom_line layer for the dataset filtered to include just Part-Time Faculty with the linewidth set to 1.5 (instead of the default .5). I think this plot accomplishes the goal of emphasizing the growth in Part-Time Faculty.

```{r plot1edited, fig.width=10, fig.height=6}
staff_long %>%
  ggplot(aes(
    x = year,
    y = value,
    group = faculty_type,
    color = faculty_type
  )) +
  geom_line() +
  geom_line(data = staff_long %>% filter(faculty_type == "Part-Time Faculty"),
          aes(x = year, y = value),
          linewidth = 1.5) + 
  labs(title = "Instructional Staff Employees, 1975-2011",
       x = "Year", y = "Percentage of Hires",
       color = "Faculty Type") +
  scale_y_continuous(labels = label_percent(scale = 1)) +
  theme_minimal()
```

## Fisheries
### Exercise 3
Can you help the researcher improve the data visualizations he made from the Fisheries and Aquaculture Department of the Food and Agriculture Organization of the United Nations' (FAO) data on the fisheries production of different countries? First, brainstorm how you would improve it. Then create the improved visualization and document your changes/decisions with bullet points. It’s ok if some of your improvements are aspirational, i.e. you don’t know how to implement it, but you think it’s a good idea. Implement what you can and leave notes identifying the aspirational improvements that could not be made. (You don’t need to recreate their plots in order to improve them)

```{r fishdata}
fisheries <- read_csv("data/fisheries.csv")
```

• I have no clue what I'm looking at with the first graph——the one with countries on the x-axis and (what I only know from your description to be) total harvest on the y-axis (there are no anxis labels). 