---
title: "Lab 06 - Ugly charts and Simpson's paradox"
author: "Noah Booker"
date: "3/28/25"
output: github_document
---
# Take a Sad Plot and Make It Better

## Instructional staff employment trends
### Load packages and data

```{r load-packages, message = FALSE}
library(tidyverse) 
library(scales)
#library(dsbox)
#library(mosaicData) 
```


```{r codefromlabintro}
staff <- read_csv("data/instructional-staff.csv")

staff_long <- staff %>%
  pivot_longer(cols = -faculty_type, names_to = "year") %>%
  mutate(value = as.numeric(value))

staff_long %>%
  ggplot(aes(x = year, y = value, color = faculty_type)) +
  geom_line()

staff_long %>%
  ggplot(aes(
    x = year,
    y = value,
    group = faculty_type,
    color = faculty_type
  )) +
  geom_line()
```

### Exercise 1

Include the line plot you made above in your report and make sure the figure width is large enough to make it legible.
Also fix the title, axis labels, and legend label.

```{r plot1, fig.width=10, fig.height=6}
staff_long %>%
  ggplot(aes(
    x = year,
    y = value,
    group = faculty_type,
    color = faculty_type
  )) +
  geom_line() +
  labs(title = "Instructional Staff Employees, 1975-2011",
       x = "Year", y = "Percentage of Hires",
       color = "Faculty Type") +
  scale_y_continuous(labels = label_percent(scale = 1)) +
  theme_minimal()
```


### Exercise 2

Suppose the objective of this plot was to show that the proportion of part-time faculty have gone up over time compared to other instructional staff types. What changes would you propose making to this plot to tell this story?

The current plot seems to tell that story pretty well. The only line that shows a consistent positive slope over the years is Part-Time Faculty. But perhaps I'm overlooking some changes that could make that fact more obvious...

Solution: I had the idea of trying to make the line for Part-Time Faculty thicker than the other lines, and asked Claude.ai for a way to do this. Claude helped me to create the code below which adds in a separate geom_line layer for the dataset filtered to include just Part-Time Faculty with the linewidth set to 1.5 (instead of the default .5). I think this plot accomplishes the goal of emphasizing the growth in Part-Time Faculty.

```{r plot1edited, fig.width=10, fig.height=6}
staff_long %>%
  ggplot(aes(
    x = year,
    y = value,
    group = faculty_type,
    color = faculty_type
  )) +
  geom_line() +
  geom_line(data = staff_long %>% filter(faculty_type == "Part-Time Faculty"),
          aes(x = year, y = value),
          linewidth = 1.5) + 
  labs(title = "Instructional Staff Employees, 1975-2011",
       x = "Year", y = "Percentage of Hires",
       color = "Faculty Type") +
  scale_y_continuous(labels = label_percent(scale = 1)) +
  theme_minimal()
```

## Fisheries
### Exercise 3
Can you help the researcher improve the data visualizations he made from the FOA data on the fisheries production of different countries? First, brainstorm how you would improve it. Then create the improved visualization and document your changes/decisions with bullet points. It’s ok if some of your improvements are aspirational, i.e. you don’t know how to implement it, but you think it’s a good idea. Implement what you can and leave notes identifying the aspirational improvements that could not be made. (You don’t need to recreate their plots in order to improve them)

```{r fishdata}
fisheries <- read_csv("data/fisheries.csv")
```

#### Brainstorm
I have no clue what I'm looking at with the first graph——the one with countries on the x-axis and (what I only know from your description to be) total harvest on the y-axis (there are no anxis labels). There's a blue curve and a red curve that both skyrocket on the left side (which I'm assuming shows that China has a lot of harvest since they're the first contry on the x-axis). But then there's little peaks as you move righward along the curves, but the peaks don't seem to correspond to where the other countries are located on the x-axis. So, I have no clue.
The other two visualizations are 3D pie charts, one showing each country's proportion of the total capture tonnage, and the other showing each country's proportion of the total aquaculture tonnage. I learned from Module 5 that bar charts are bad and 3D is bad.
The description in the lab says that the visualization excludes countries with less than 100,000 tons. This is probably a good call since there are 216 countries in the dataset.
It seems to me like a reasonable primary goal with creating a visualization for this data is to be able to compare countries' total tonnage of fish. Second, in visualizing tonnage of fish by country, we might also want to display how much of each country's total tonnage came from capture or aquaculture. Third, ignoring total tonnage, we may be interested in comparing each country by the proportion of their fish which came from capture or aquaculture.

#### Improved Visualization
• Let's first get an idea of how many countries we can usefully compare. The lab description says that the visualization excludes countries with less than 100,000 tons. Let's see how many countries that leaves us with.

```{r tonnage_cutoff}
#fisheries %>% 
  #select(total)
```
• A barplot seems to me like the most appropriate way to compare countries on their fishery production (much better than a pie chart or whatever that other graph is).

```{r barplot1}
```