---
title: "Lab 06 - Ugly charts and Simpson's paradox"
author: "Noah Booker"
date: "3/28/25"
output: github_document
---

### Load packages and data

```{r load-packages, message = FALSE}
library(tidyverse) 
library(scales)
#library(dsbox)
#library(mosaicData) 
```

```{r codefromlabintro}
staff <- read_csv("data/instructional-staff.csv")

staff_long <- staff %>%
  pivot_longer(cols = -faculty_type, names_to = "year") %>%
  mutate(value = as.numeric(value))

staff_long %>%
  ggplot(aes(x = year, y = value, color = faculty_type)) +
  geom_line()

staff_long %>%
  ggplot(aes(
    x = year,
    y = value,
    group = faculty_type,
    color = faculty_type
  )) +
  geom_line()
```

### Exercise 1

Include the line plot you made above in your report and make sure the figure width is large enough to make it legible.
Also fix the title, axis labels, and legend label.

```{r plot1, fig.width=10, fig.height=6}
staff_long %>%
  ggplot(aes(
    x = year,
    y = value,
    group = faculty_type,
    color = faculty_type
  )) +
  geom_line() +
  labs(title = "Instructional Staff Employees, 1975-2011",
       x = "Year", y = "Percentage of Hires",
       color = "Faculty Type") +
  scale_y_continuous(labels = label_percent(scale = 1)) +
  theme_minimal()
```


### Exercise 2

Suppose the objective of this plot was to show that the proportion of part-time faculty have gone up over time compared to other instructional staff types. What changes would you propose making to this plot to tell this story?

The current plot seems to tell that story pretty well. The only line that shows a consistent positive slope over the years is Part-Time Faculty. But perhaps I'm overlooking some changes that could make that fact more obvious...

Solution: I had the idea of trying to make the line for Part-Time Faculty thicker than the other lines, and asked Claude.ai for a way to do this. Claude helped me to create the code below which adds in a separate geom_line layer for the dataset filtered to include just Part-Time Faculty with the linewidth set to 1.5 (instead of the default .5). I think this plot accomplishes the goal of emphasizing the growth in Part-Time Faculty.

```{r plot1edited, fig.width=10, fig.height=6}
staff_long %>%
  ggplot(aes(
    x = year,
    y = value,
    group = faculty_type,
    color = faculty_type
  )) +
  geom_line() +
  geom_line(data = staff_long %>% filter(faculty_type == "Part-Time Faculty"),
          aes(x = year, y = value),
          linewidth = 1.5) + 
  labs(title = "Instructional Staff Employees, 1975-2011",
       x = "Year", y = "Percentage of Hires",
       color = "Faculty Type") +
  scale_y_continuous(labels = label_percent(scale = 1)) +
  theme_minimal()
```
Remove this text, and add your answer for Exercise 1 here.
Add code chunks as needed.
Don't forget to label your code chunk.
Do not use spaces in code chunk labels.

### Exercise 3

...

Add exercise headings as needed.
