---
title: "Lab 06 - Ugly charts and Simpson's paradox"
author: "Noah Booker"
date: "3/29/25"
output: github_document
---
# Take a Sad Plot and Make It Better

## Instructional staff employment trends
### Load packages and data

```{r load-packages, message = FALSE}
library(tidyverse) 
library(scales)
#library(dsbox)
#library(mosaicData) 
```


```{r codefromlabintro}
staff <- read_csv("data/instructional-staff.csv")

staff_long <- staff %>%
  pivot_longer(cols = -faculty_type, names_to = "year") %>%
  mutate(value = as.numeric(value))

staff_long %>%
  ggplot(aes(x = year, y = value, color = faculty_type)) +
  geom_line()

staff_long %>%
  ggplot(aes(
    x = year,
    y = value,
    group = faculty_type,
    color = faculty_type
  )) +
  geom_line()
```

### Exercise 1

Include the line plot you made above in your report and make sure the figure width is large enough to make it legible.
Also fix the title, axis labels, and legend label.

```{r plot1, fig.width=10, fig.height=6}
staff_long %>%
  ggplot(aes(
    x = year,
    y = value,
    group = faculty_type,
    color = faculty_type
  )) +
  geom_line() +
  labs(title = "Instructional Staff Employees, 1975-2011",
       x = "Year", y = "Percentage of Hires",
       color = "Faculty Type") +
  scale_y_continuous(labels = label_percent(scale = 1)) +
  theme_minimal()
```


### Exercise 2

Suppose the objective of this plot was to show that the proportion of part-time faculty have gone up over time compared to other instructional staff types. What changes would you propose making to this plot to tell this story?

The current plot seems to tell that story pretty well. The only line that shows a consistent positive slope over the years is Part-Time Faculty. But perhaps I'm overlooking some changes that could make that fact more obvious...

Solution: I had the idea of trying to make the line for Part-Time Faculty thicker than the other lines, and asked Claude.ai for a way to do this. Claude helped me to create the code below which adds in a separate geom_line layer for the dataset filtered to include just Part-Time Faculty with the linewidth set to 1.5 (instead of the default .5). I think this plot accomplishes the goal of emphasizing the growth in Part-Time Faculty.

```{r plot1_2, fig.width=10, fig.height=6}
staff_long %>%
  ggplot(aes(
    x = year,
    y = value,
    group = faculty_type,
    color = faculty_type
  )) +
  geom_line() +
  geom_line(data = staff_long %>% filter(faculty_type == "Part-Time Faculty"),
          aes(x = year, y = value),
          linewidth = 1.5) + 
  labs(title = "Instructional Staff Employees, 1975-2011",
       x = "Year", y = "Percentage of Hires",
       color = "Faculty Type") +
  scale_y_continuous(labels = label_percent(scale = 1)) +
  theme_minimal()
```

## Fisheries
### Exercise 3
Can you help the researcher improve the data visualizations he made from the FOA data on the fisheries production of different countries? First, brainstorm how you would improve it. Then create the improved visualization and document your changes/decisions with bullet points. It’s ok if some of your improvements are aspirational, i.e. you don’t know how to implement it, but you think it’s a good idea. Implement what you can and leave notes identifying the aspirational improvements that could not be made. (You don’t need to recreate their plots in order to improve them.)

```{r fishdata}
fisheries <- read_csv("data/fisheries.csv")
```

#### Brainstorm
I have no clue what I'm looking at with the first graph——the one with countries on the x-axis and (what I only know from your description to be) total harvest on the y-axis (there are no anxis labels). There's a blue curve and a red curve that both skyrocket on the left side (which I'm assuming shows that China has a lot of harvest since they're the first contry on the x-axis). But then there's little peaks as you move righward along the curves, but the peaks don't seem to correspond to where the other countries are located on the x-axis. So, I have no clue.
The other two visualizations are 3D pie charts, one showing each country's proportion of the total capture tonnage, and the other showing each country's proportion of the total aquaculture tonnage. I learned from Module 5 that bar charts are bad and 3D is bad.
The description in the lab says that the visualization excludes countries with less than 100,000 tons. This is probably a good call since there are 216 countries in the dataset.
It seems to me like a reasonable primary goal with creating a visualization for this data is to be able to compare countries' total tonnage of fish. Second, in visualizing tonnage of fish by country, we might also want to display how much of each country's total tonnage came from capture or aquaculture. Third, ignoring total tonnage, we may be interested in comparing each country by the proportion of their fish which came from capture or aquaculture.

#### Improved Visualization
• Let's first get an idea of how many countries we can usefully compare. The lab description says that the visualization excludes countries with less than 100,000 tons. Let's see how many countries that leaves us with.

```{r tonnage_cutoff}
fisheries %>% 
  filter(total >= 100000)
```

• Making the cutoff at 100,000 tons still leaves us with a dataframe with 82 countries. That seems like too many to usefully compare. I think it seems more reasonable to compare the top ten.

```{r topten}
fisheries_top10 <- fisheries %>% 
  arrange(desc(total)) %>% 
  slice(1:10) #Thanks to Claude for suggesting the slice( ) function.
```

• A barplot seems to me like the most appropriate way to compare countries on their fishery production (much better than a pie chart or whatever that other graph is).

```{r plot2}
fisheries_top10 %>% 
  ggplot(aes(x = country, y = total)) +
  geom_col()
```

• Order matters: order countries on x-axis in order of fish production.
```{r reorder_countries}
fisheries_top10 <- fisheries_top10 %>% 
  mutate(
    country = fct_relevel(
      country,
      "China", "Indonesia", "India", "Vietnam", "United States", 
      "Russia", "Japan", "Philippines", "Peru", "Bangladesh"))
```
• Use meaningful and nonredundant labels: Improve y-axis label, remove x-axis label, add title and subtitle.

• Make country labels angled for easier reading (I like this option better than coord_flip).

• Fix y-axis value labels to be regular numbers, in increments of 10 million.
```{r plot2_2}
fisheries_top10 %>% 
  ggplot(aes(x = country, y = total)) +
  geom_col() +
  labs(x = NULL, y = "Tons (Millions)", title = "Top 10 World Fisheries' Production, 2016",
       subtitle = "Data from the Fisheries and Aquaculture Department of the FOA of the United Nations") +
  theme_minimal() +
  #I recycled this code for angling the x-axis value labels from the Brexit activity. (Claude helped create it.)
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)) +
  #Claude helped with the code below to get the y-axis value labels how I wanted them.
  scale_y_continuous(
    breaks = seq(0, max(fisheries_top10$total), by = 10000000),
    labels = function(x) x/1000000
  )
```

• Show each country's total's composition of captured vs. farmed fish.

```{r plot2_3}
# Reshape the data from wide to long format
fisheries_top10_long <- fisheries_top10 %>%
  pivot_longer(cols = c(capture, aquaculture), 
               names_to = "production_type",
               values_to = "value")

# Create the stacked bar chart
fisheries_top10_long %>%
  ggplot(aes(x = country, y = value, fill = production_type)) +
  geom_col() +
  labs(x = NULL, 
       y = "Tons (Millions)", 
       title = "Top 10 World Fisheries' Production, 2016",
       subtitle = "Data from the Fisheries and Aquaculture Department of the FOA of the United Nations",
       fill = "Production Type") +
  theme_minimal() +
  #Thanks Claude for helping to adjust the legend position.
  theme(
  axis.text.x = element_text(angle = 45, hjust = 1),
  legend.position = "inside",
  legend.position.inside = c(0.85, 0.85),
  legend.background = element_rect(fill = "white", color = NA, linewidth = 0.5),
  legend.margin = margin(6, 6, 6, 6)
) +
  scale_y_continuous(
    breaks = seq(0, max(fisheries_top10$total), by = 5000000),
    labels = function(x) x/1000000
  ) + #Used the colors from the Brexit activity.
  scale_fill_manual(values = c(
    "aquaculture" = "#ef8a62",
    "capture" = "#67a9cf"),
labels = c(
  "aquaculture" = "Aquaculture",
  "capture" = "Capture"
))
```

#### Other things I would do with more time

If I had more time to spend on this lab I might do some other things.

• Compare the top ten countries on their proportion of fishery production that comes from aquaculture or capture while ignoring total producton (similar to how we compared to proportion of opinions by region in the Brexit data, ignoring differences in regional sample sizes).

• Create a variable that represents each countries proportion of the overall total fishery production and compare the top 10 countries. This would tell us what proportion of the world's total fishery production China, for example, accounts for. The pie charts shown in the lab module sort of do this (separately for capture and aquaculture), but a bar graph would make it easier to quantify.