---
title: "Lab 06 - Ugly charts and Simpson's paradox"
author: "Noah Booker"
date: "3/28/25"
output: github_document
---

### Load packages and data

```{r load-packages, message = FALSE}
library(tidyverse) 
#library(dsbox)
#library(mosaicData) 
```

```{r}
staff <- read_csv("data/instructional-staff.csv")

staff_long <- staff %>%
  pivot_longer(cols = -faculty_type, names_to = "year") %>%
  mutate(value = as.numeric(value))

staff_long %>%
  ggplot(aes(x = year, y = value, color = faculty_type)) +
  geom_line()

staff_long %>%
  ggplot(aes(
    x = year,
    y = value,
    group = faculty_type,
    color = faculty_type
  )) +
  geom_line()
```

### Exercise 1

Include the line plot you made above in your report and make sure the figure width is large enough to make it legible.
Also fix the title, axis labels, and legend label.

Remove this text, and add your answer for Exercise 1 here.
Add code chunks as needed.
Don't forget to label your code chunk.
Do not use spaces in code chunk labels.

### Exercise 2

Remove this text, and add your answer for Exercise 1 here.
Add code chunks as needed.
Don't forget to label your code chunk.
Do not use spaces in code chunk labels.

### Exercise 3

...

Add exercise headings as needed.
